The formal verification of the ctm approach to forcing \thanks{Supported by Secyt-UNC projects 33620180100855CB and 33620180100465CB, and Conicet.}

### 2023-09-28 Independence_CH/forcing.tex
First typo spotted after arXiv v1
Explanation of separation, strong_repl, blowup, ack Commelin

For stating the axiom schemes, \session{ZF-Constructible} defines
the expressions
\[
  \isa{separation}(N,Q)
  \text{ and }
  \isa{strong{\uscore}replacement}(N,R)
\]
whose first argument $N$ is a class and their second arguments $Q$ and
$R$ are predicates of types $R :: \tyi \fun \tyo$ and $Q ::
\tyi \fun \tyi \fun \tyo$, respectively. The Separation Axiom appears in
\locale{M{\uscore}Z{\uscore}basic} as follows, where 
\isa{\isasymphi} is free and \isa{\isacharat} denotes
list concatenation:
\begin{isabelle}
separation{\isacharunderscore}{\kern0pt}ax{\isacharcolon}{\kern0pt}\ {\isachardoublequoteopen}{\isasymphi}\ {\isasymin}\ formula\ {\isasymLongrightarrow}\ env\ {\isasymin}\ list{\isacharparenleft}{\kern0pt}M{\isacharparenright}{\kern0pt}\ {\isasymLongrightarrow}\isanewline
\ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ arity{\isacharparenleft}{\kern0pt}{\isasymphi}{\isacharparenright}{\kern0pt}\ {\isasymle}\ {\isadigit{1}}\ {\isacharplus}{\kern0pt}\isactrlsub {\isasymomega}\ length{\isacharparenleft}{\kern0pt}env{\isacharparenright}{\kern0pt}\ {\isasymLongrightarrow}\isanewline
\ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ separation{\isacharparenleft}{\kern0pt}{\isacharhash}{\kern0pt}{\isacharhash}{\kern0pt}M{\isacharcomma}{\kern0pt}{\isasymlambda}x{\isachardot}{\kern0pt}\ {\isacharparenleft}{\kern0pt}M{\isacharcomma}{\kern0pt}\ {\isacharbrackleft}{\kern0pt}x{\isacharbrackright}{\kern0pt}\ {\isacharat}{\kern0pt}\ env\ {\isasymTurnstile}\ {\isasymphi}{\isacharparenright}{\kern0pt}{\isacharparenright}{\kern0pt}{\isachardoublequoteclose}
\end{isabelle}
%% \sout{we can only show
%% that this form of the axioms hold in generic extensions.}
Note that the predicate $Q$ mentioned above corresponds to the
satisfaction of \isa{\isasymphi}.

In contrast to Separation, we stated each instance of the Replacement
Axiom separately by means of the
$\isa{replacement{\uscore}assm}(M,\mathit{env},\phi)$ predicate:
\begin{isabelle}
{\isasymphi}\ {\isasymin}\ formula\ {\isasymlongrightarrow}\ env\ {\isasymin}\ list{\isacharparenleft}{\kern0pt}M{\isacharparenright}{\kern0pt}\ {\isasymlongrightarrow}\isanewline
\ \ arity{\isacharparenleft}{\kern0pt}{\isasymphi}{\isacharparenright}{\kern0pt}\ {\isasymle}\ {\isadigit{2}}\ {\isacharplus}{\kern0pt}\isactrlsub {\isasymomega}\ length{\isacharparenleft}{\kern0pt}env{\isacharparenright}{\kern0pt}\ {\isasymlongrightarrow}\isanewline
\ \ \ \ strong{\isacharunderscore}{\kern0pt}replacement{\isacharparenleft}{\kern0pt}{\isacharhash}{\kern0pt}{\isacharhash}{\kern0pt}M{\isacharcomma}{\kern0pt}{\isasymlambda}x\ y{\isachardot}{\kern0pt}\ {\isacharparenleft}{\kern0pt}M\ {\isacharcomma}{\kern0pt}\ {\isacharbrackleft}{\kern0pt}x{\isacharcomma}{\kern0pt}y{\isacharbrackright}{\kern0pt}{\isacharat}{\kern0pt}env\ {\isasymTurnstile}\ {\isasymphi}{\isacharparenright}{\kern0pt}{\isacharparenright}{\kern0pt}{\isachardoublequoteclose}
\end{isabelle}
%% % Same as above
%% \begin{multline}\label{eq:replacement_assm_def}
%% \varphi \in \formula  \limp \mathit{env} \in \isa{list}(M) \limp \isa{arity}(\varphi) \leq 2+ \isa{length}(\mathit{env}) \limp \\
%%  \isa{strong{\uscore}replacement}(\isa{\#\#} M, \lambda x\, y.\ (M, [x,y]
%% \mathbin{@} \mathit{env}  \models \varphi))
%% \end{multline}
In turn, the \isa{{\isasymcdot}Replacement{\isasymcdot}} function
takes a formula code and returns the corresponding replacement
instance:
\begin{isabelle}
{\isasymphi}\ {\isasymin}\ formula\ {\isasymLongrightarrow}\isanewline
{\isacharparenleft}{\kern0pt}M{\isacharcomma}{\kern0pt}\ {\isacharbrackleft}{\kern0pt}{\isacharbrackright}{\kern0pt}\ {\isasymTurnstile}\ {\isasymcdot}Replacement{\isacharparenleft}{\kern0pt}{\isasymphi}{\isacharparenright}{\kern0pt}{\isasymcdot}{\isacharparenright}{\kern0pt}\ {\isasymlongleftrightarrow}\ {\isacharparenleft}{\kern0pt}{\isasymforall}env{\isachardot}{\kern0pt}\ replacement{\isacharunderscore}{\kern0pt}assm{\isacharparenleft}{\kern0pt}M{\isacharcomma}{\kern0pt}\ env{\isacharcomma}{\kern0pt}\ {\isasymphi}{\isacharparenright}{\kern0pt}{\isacharparenright}
\end{isabelle}
%%%%%%%
%% \sout{In \session{ZF-Constructible} some results are proved assuming some instances of
%% separation or replacement (mejor si ponemos ejemplo concreto). In contrast,
%% our locale \isa{M{\uscore}Z{\uscore}basic} assumes the full scheme of separation in terms
%% of satisfaction of some internalized formula but no instance of replacement. (Mostrar el enunciado de separation\_ax:
%% es distinto a Reemplazo y se nombra después)
%% 
%% In order to have a finer control of the instances of replacement needed,
%% we postulate further locales with the instances needed to interpret at
%% the class\dots
%% (tal vez no sea necesario mostrar la definición de .Replacement.)}
%%%%%%%

>>>>

For stating the axiom schemes, \session{ZF-Constructible} defines
$\isa{separation}(N,Q)$:
\[
  \forall z[N].\ \exists y[N].\ \forall
  x[N].\ x\in y \longleftrightarrow x\in z \land Q(x),
\]
and $\isa{strong{\uscore}replacement}(N,R)$:
\begin{multline*}
  \forall
  A[N].\ \isa{univalent}(N,A,R) \longrightarrow \\
  (\exists Y[N].\ \forall b[N].\ b\in Y\longleftrightarrow (\exists x[N].\ x\in
  A \land R(x,b))),
\end{multline*}
%% \begin{isabelle}%
%% separation{\isacharparenleft}{\kern0pt}N{\isacharcomma}{\kern0pt}\ Q{\isacharparenright}{\kern0pt}\ {\isasymequiv}\ {\isasymforall}z{\isacharbrackleft}{\kern0pt}N{\isacharbrackright}{\kern0pt}{\isachardot}{\kern0pt}\ {\isasymexists}y{\isacharbrackleft}{\kern0pt}N{\isacharbrackright}{\kern0pt}{\isachardot}{\kern0pt}\ {\isasymforall}x{\isacharbrackleft}{\kern0pt}N{\isacharbrackright}{\kern0pt}{\isachardot}{\kern0pt}\ x\ {\isasymin}\ y\ {\isasymlongleftrightarrow}\ x\ {\isasymin}\ z\ {\isasymand}\ Q{\isacharparenleft}{\kern0pt}x{\isacharparenright}{\kern0pt}%
%% \end{isabelle}%
%% \begin{isabelle}%
%% strong{\isacharunderscore}{\kern0pt}replacement{\isacharparenleft}{\kern0pt}N{\isacharcomma}{\kern0pt}\ R{\isacharparenright}{\kern0pt}\ {\isasymequiv}\isanewline
%% {\isasymforall}A{\isacharbrackleft}{\kern0pt}N{\isacharbrackright}{\kern0pt}{\isachardot}{\kern0pt}\isanewline
%% \isaindent{\ \ \ }univalent{\isacharparenleft}{\kern0pt}N{\isacharcomma}{\kern0pt}\ A{\isacharcomma}{\kern0pt}\ R{\isacharparenright}{\kern0pt}\ {\isasymlongrightarrow}\ {\isacharparenleft}{\kern0pt}{\isasymexists}Y{\isacharbrackleft}{\kern0pt}N{\isacharbrackright}{\kern0pt}{\isachardot}{\kern0pt}\ {\isasymforall}b{\isacharbrackleft}{\kern0pt}N{\isacharbrackright}{\kern0pt}{\isachardot}{\kern0pt}\ b\ {\isasymin}\ Y\ {\isasymlongleftrightarrow}\isanewline
%% \ \ \ \ \ \ {\isacharparenleft}{\kern0pt}{\isasymexists}x{\isacharbrackleft}{\kern0pt}N{\isacharbrackright}{\kern0pt}{\isachardot}{\kern0pt}\ x\ {\isasymin}\ A\ {\isasymand}
%% R{\isacharparenleft}{\kern0pt}x{\isacharcomma}{\kern0pt}\ b{\isacharparenright}{\kern0pt}{\isacharparenright}{\kern0pt}{\isacharparenright}{\kern0pt}%
%% \end{isabelle}%
whose first argument $N$ is a class and their second arguments $Q$ and
$R$ are predicates of types $Q :: \tyi \fun \tyo$ and $R ::
\tyi \fun \tyi \fun \tyo$, respectively (the \isatt{univalent}
predicate states that $R$ is functional in its first variable; see Appendix~\ref{sec:def-main-relative}). The Separation Axiom appears in
\locale{M{\uscore}Z{\uscore}basic} as follows, where 
\isa{\isasymphi} is free and \isa{\isacharat} denotes
list concatenation:
\begin{isabelle}
separation{\isacharunderscore}{\kern0pt}ax{\isacharcolon}{\kern0pt}\ {\isachardoublequoteopen}{\isasymphi}\ {\isasymin}\ formula\ {\isasymLongrightarrow}\ env\ {\isasymin}\ list{\isacharparenleft}{\kern0pt}M{\isacharparenright}{\kern0pt}\ {\isasymLongrightarrow}\isanewline
\ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ arity{\isacharparenleft}{\kern0pt}{\isasymphi}{\isacharparenright}{\kern0pt}\ {\isasymle}\ {\isadigit{1}}\ {\isacharplus}{\kern0pt}\isactrlsub {\isasymomega}\ length{\isacharparenleft}{\kern0pt}env{\isacharparenright}{\kern0pt}\ {\isasymLongrightarrow}\isanewline
\ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ separation{\isacharparenleft}{\kern0pt}{\isacharhash}{\kern0pt}{\isacharhash}{\kern0pt}M{\isacharcomma}{\kern0pt}{\isasymlambda}x{\isachardot}{\kern0pt}\ {\isacharparenleft}{\kern0pt}M{\isacharcomma}{\kern0pt}\ {\isacharbrackleft}{\kern0pt}x{\isacharbrackright}{\kern0pt}\ {\isacharat}{\kern0pt}\ env\ {\isasymTurnstile}\ {\isasymphi}{\isacharparenright}{\kern0pt}{\isacharparenright}{\kern0pt}{\isachardoublequoteclose}
\end{isabelle}
%% \sout{we can only show
%% that this form of the axioms hold in generic extensions.}
Note that the predicate $Q$ mentioned above corresponds to the
satisfaction of \isa{\isasymphi}.

In contrast to Separation, we stated each instance of the Replacement
Axiom separately by means of the
$\isa{replacement{\uscore}assm}(M,\mathit{env},\phi)$ predicate:
\begin{isabelle}
{\isasymphi}\ {\isasymin}\ formula\ {\isasymlongrightarrow}\ env\ {\isasymin}\ list{\isacharparenleft}{\kern0pt}M{\isacharparenright}{\kern0pt}\ {\isasymlongrightarrow}\isanewline
\ \ arity{\isacharparenleft}{\kern0pt}{\isasymphi}{\isacharparenright}{\kern0pt}\ {\isasymle}\ {\isadigit{2}}\ {\isacharplus}{\kern0pt}\isactrlsub {\isasymomega}\ length{\isacharparenleft}{\kern0pt}env{\isacharparenright}{\kern0pt}\ {\isasymlongrightarrow}\isanewline
\ \ \ \ strong{\isacharunderscore}{\kern0pt}replacement{\isacharparenleft}{\kern0pt}{\isacharhash}{\kern0pt}{\isacharhash}{\kern0pt}M{\isacharcomma}{\kern0pt}{\isasymlambda}x\ y{\isachardot}{\kern0pt}\ {\isacharparenleft}{\kern0pt}M\ {\isacharcomma}{\kern0pt}\ {\isacharbrackleft}{\kern0pt}x{\isacharcomma}{\kern0pt}y{\isacharbrackright}{\kern0pt}{\isacharat}{\kern0pt}env\ {\isasymTurnstile}\ {\isasymphi}{\isacharparenright}{\kern0pt}{\isacharparenright}{\kern0pt}{\isachardoublequoteclose},
\end{isabelle}
%% % Same as above
%% \begin{multline}\label{eq:replacement_assm_def}
%% \varphi \in \formula  \limp \mathit{env} \in \isa{list}(M) \limp \isa{arity}(\varphi) \leq 2+ \isa{length}(\mathit{env}) \limp \\
%%  \isa{strong{\uscore}replacement}(\isa{\#\#} M, \lambda x\, y.\ (M, [x,y]
%% \mathbin{@} \mathit{env}  \models \varphi))
%% \end{multline}
which encodes the instance of replacement for the model $M$
corresponding to the predicate $R(x,y)$ given by
$\phi^M(x,y,x_0,\dots,x_n)$, where $\isatt{env} = [x_0,\dots,x_n]$.

In turn, the \isa{{\isasymcdot}Replacement{\isasymcdot}} function
takes a formula code and returns the corresponding replacement
instance:
\begin{isabelle}
{\isasymphi}\ {\isasymin}\ formula\ {\isasymLongrightarrow}\isanewline
{\isacharparenleft}{\kern0pt}M{\isacharcomma}{\kern0pt}\ {\isacharbrackleft}{\kern0pt}{\isacharbrackright}{\kern0pt}\ {\isasymTurnstile}\ {\isasymcdot}Replacement{\isacharparenleft}{\kern0pt}{\isasymphi}{\isacharparenright}{\kern0pt}{\isasymcdot}{\isacharparenright}{\kern0pt}\ {\isasymlongleftrightarrow}\ {\isacharparenleft}{\kern0pt}{\isasymforall}env{\isachardot}{\kern0pt}\ replacement{\isacharunderscore}{\kern0pt}assm{\isacharparenleft}{\kern0pt}M{\isacharcomma}{\kern0pt}\ env{\isacharcomma}{\kern0pt}\ {\isasymphi}{\isacharparenright}{\kern0pt}{\isacharparenright}
\end{isabelle}
%%%%%%%
%% \sout{In \session{ZF-Constructible} some results are proved assuming some instances of
%% separation or replacement (mejor si ponemos ejemplo concreto). In contrast,
%% our locale \isa{M{\uscore}Z{\uscore}basic} assumes the full scheme of separation in terms
%% of satisfaction of some internalized formula but no instance of replacement. (Mostrar el enunciado de separation\_ax:
%% es distinto a Reemplazo y se nombra después)
%% 
%% In order to have a finer control of the instances of replacement needed,
%% we postulate further locales with the instances needed to interpret at
%% the class\dots
%% (tal vez no sea necesario mostrar la definición de .Replacement.)}
%%%%%%%
